---
title: "Car weight and fuel efficiency"
author: "Rupesh Patel"
date: 2026-09-21
execute:
  echo: true
---

## Introduction

Do heavier cars tend to have lower fuel efficiency? In this post, I
explore the relationship between car weight and miles per gallon.
I also compare average fuel efficiency across cylinder groups.

A higher miles-per-gallon value means the car can travel farther
using the same amount of fuel.

## Data source

I use the built-in R dataset
[mtcars: Motor Trend Car Road Tests](https://stat.ethz.ch/R-manual/R-devel/library/datasets/html/mtcars.html).

The data was extracted from the 1974 Motor Trend magazine and contains
32 cars from the 1973–74 model years. The R documentation cites
[Henderson and Velleman (1981)](https://doi.org/10.2307/2530428).

I use the copy included with R, so no separate data download is needed.

## Set up

```{r}
#| label: r-setup
#| message: false
#| warning: false

library(dplyr)
library(ggplot2)
```

The setup chunk suppresses package startup messages and warnings
to keep the page easier to read.

## Explore the data

I select the columns needed for this analysis and check for missing
values.

```{r}
#| label: explore-cars

cars <- mtcars |>
  select(mpg, wt, cyl)

head(cars)
nrow(cars)
colSums(is.na(cars))
```

There are 32 cars and no missing values in these three columns.
The variables are:

- `mpg`: miles per US gallon.
- `wt`: car weight in thousands of pounds.
- `cyl`: number of cylinders.

## Compare fuel efficiency by cylinder group

I calculate the number of cars and the average fuel efficiency
for each cylinder group.

```{r}
#| label: summarize-cars

cylinder_summary <- cars |>
  group_by(cyl) |>
  summarise(
    number_of_cars = n(),
    average_mpg = mean(mpg),
    .groups = "drop"
  )

knitr::kable(
  cylinder_summary,
  digits = 1,
  col.names = c("Cylinders", "Number of cars", "Average MPG")
)
```

The four-cylinder cars have the highest average fuel efficiency,
at approximately 26.7 MPG. The averages for six-cylinder and
eight-cylinder cars are approximately 19.7 and 15.1 MPG.

These are comparisons between the cars in this dataset. They do
not show that cylinder count alone explains the differences.

## Plot weight against fuel efficiency

I use a scatter plot to check how fuel efficiency changes with car
weight. Each point represents one car. I treat cylinder count as a
category so each group gets a separate colour.

```{r}
#| label: fig-car-weight-mpg
#| fig-cap: "Car weight and fuel efficiency for 32 cars, coloured by cylinder count."
#| fig-width: 7
#| fig-height: 4.5

ggplot(cars, aes(x = wt, y = mpg, colour = factor(cyl))) +
  geom_point(size = 3, alpha = 0.8) +
  labs(
    title = "Car weight and fuel efficiency",
    x = "Weight (1,000 pounds)",
    y = "Fuel efficiency (miles per US gallon)",
    colour = "Cylinders"
  ) +
  theme_minimal()
```

The plot shows a downward pattern: heavier cars generally have lower
MPG values. The eight-cylinder cars also tend to appear among the
heavier cars with lower fuel efficiency.

## Measure the relationship

I calculate the Pearson correlation between weight and MPG.

```{r}
#| label: weight-mpg-correlation

weight_mpg_correlation <- cor(cars$wt, cars$mpg)

round(weight_mpg_correlation, 3)
```

The correlation is approximately -0.868, indicating a strong negative
linear relationship in this dataset.

## Conclusion

Heavier cars tend to have lower fuel efficiency in this dataset.
Four-cylinder cars also have higher average MPG than the six-cylinder
and eight-cylinder groups.

However, this is an observational comparison. Weight, cylinder count,
and other car characteristics vary together, so the analysis does not
isolate the effect of weight alone.

The data contains only 32 older cars. These results should not be
assumed to describe modern vehicles.