---
title: "Comparing fuel efficiency using R and Python"
author: "Rupesh Patel"
date: 2026-09-21
engine: knitr
execute:
  echo: true
---

## Introduction

In this post, I use R and Python together to compare fuel efficiency
between automatic and manual cars.

I prepare the data in R, calculate group summaries in Python, and
bring the results back into R to create a table and plot. The
reticulate package allows the two languages to share objects without
saving intermediate CSV files.

## Data source

I use the built-in R dataset
[mtcars: Motor Trend Car Road Tests](https://stat.ethz.ch/R-manual/R-devel/library/datasets/html/mtcars.html).

The data was extracted from the 1974 Motor Trend magazine and contains
32 cars from the 1973–74 model years. The R documentation cites
[Henderson and Velleman (1981)](https://doi.org/10.2307/2530428).

The dataset comes with R, so no separate download is needed.

## Set up R and Python

```{r}
#| label: bonus-setup
#| message: false
#| warning: false

library(reticulate)
library(dplyr)
library(ggplot2)

use_virtualenv("../../.venv", required = TRUE)

py_config()
```

This post is inside `posts/r-and-python`, so `../../.venv` points to
the Python environment at the repository root. This relative path
also works when the repository is cloned into a different location.

The configuration output shows which Python interpreter is being used.

## Prepare the data in R

I select fuel efficiency, weight, and transmission type. The original
transmission variable uses 0 for automatic and 1 for manual, so I
create readable labels.

```{r}
#| label: prepare-bonus-data

cars_bonus <- mtcars |>
  select(mpg, wt, am) |>
  mutate(
    transmission = if_else(am == 0, "Automatic", "Manual")
  )

head(cars_bonus)
colSums(is.na(cars_bonus))
```

There are no missing values in these selected columns. Fuel efficiency
is measured in miles per US gallon, and weight is measured in
thousands of pounds.

## Pass the R data to Python

In a Python chunk, `r.cars_bonus` accesses the object created in R.
Reticulate converts the R data frame into a pandas DataFrame.

```{python}
#| label: receive-r-data

import pandas as pd

cars_python = r.cars_bonus.copy()

print(type(cars_python))
print("Number of cars:", len(cars_python))
print(cars_python.head())
```

This confirms that Python received the data prepared in R.

## Calculate the summaries in Python

I group the cars by transmission type and calculate the number of
cars, average fuel efficiency, and average weight.

```{python}
#| label: summarize-in-python

transmission_summary = (
    cars_python
    .groupby("transmission", as_index=False)
    .agg(
        number_of_cars=("mpg", "size"),
        average_mpg=("mpg", "mean"),
        average_weight=("wt", "mean")
    )
)

print(transmission_summary.round(2).to_string(index=False))
```

The manual group has an average fuel efficiency of about 24.4 MPG,
compared with 17.1 MPG for the automatic group.

However, the manual cars are also lighter on average. Their average
weight is about 2.41 thousand pounds, compared with 3.77 thousand
pounds for the automatic cars.

## Bring the Python results back to R

In an R chunk, `py$transmission_summary` accesses the summary created
in Python. I store it as an R object and display it as a table.

```{r}
#| label: receive-python-summary

summary_r <- py$transmission_summary

knitr::kable(
  summary_r,
  digits = 2,
  col.names = c(
    "Transmission",
    "Number of cars",
    "Average MPG",
    "Average weight (1,000 pounds)"
  )
)
```

The numbers in this table were calculated in Python and then
passed back to R.

## Plot the results in R

```{r}
#| label: fig-bonus-transmission-mpg
#| fig-cap: "Average fuel efficiency by transmission type. The averages were calculated in Python and plotted in R."
#| fig-width: 7
#| fig-height: 4.5

ggplot(
  summary_r,
  aes(x = transmission, y = average_mpg, fill = transmission)
) +
  geom_col(width = 0.6) +
  geom_text(
    aes(label = round(average_mpg, 1)),
    vjust = -0.5
  ) +
  scale_fill_manual(
    values = c("Automatic" = "#4477AA", "Manual" = "#228833")
  ) +
  scale_y_continuous(
    expand = expansion(mult = c(0, 0.12))
  ) +
  labs(
    title = "Average fuel efficiency by transmission",
    x = "Transmission",
    y = "Average fuel efficiency (miles per US gallon)"
  ) +
  guides(fill = "none") +
  theme_minimal()
```

## Conclusion

Manual cars have higher average MPG in this dataset, but they are
also lighter on average. This comparison does not show that changing
the transmission alone would improve fuel efficiency.

The dataset contains only 32 older cars, so these results should not
be assumed to describe modern vehicles.

This post demonstrates a shared workflow: R prepares the data,
Python calculates the summaries, and R displays the results.
The objects are passed between the languages using reticulate.